---
title: "NYPD Shooting Incident Data Report"
author: "J. Wu"
date: "2024-06-12"
output: 
  # html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
## Introducation

- Data:
The dataset analyzed consists of shooting incidents in New York City from 2006 to the end of the previous calendar year. It includes detailed information on each shooting event, such as the date, location, and demographics of victims and perpetrators.

*Goals*:
Understand Patterns: Identify trends and patterns in shooting incidents.
*Analyze Demographics*: Examine the relationship between the demographics of victims and perpetrators and the locations of incidents.
*Time-Based Analysis*: Compare the patterns of incidents before and during the COVID-19 pandemic.
*Tasks*:
- Load and preprocess the data to ensure accuracy.
Conduct exploratory data analysis to understand basic patterns and trends.
- Visualize the data to highlight key insights.
- Perform statistical tests to validate observed patterns.
- Summarize findings to provide actionable insights for policy-making and interventions.


### About Data
This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity. Please refer to the attached data footnotes for additional information about this dataset.

**Metadata Updated**: April 26, 2024

for more information, please refer  to: https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic

### Step 0: Import Library

```{r}
library(lubridate)
library(tidyverse)
```

### Step 1: Load Data
To begin with, read the dataset directlty from web with built-in funcation `read.csv()`.


```{r}

#reading data from csv file
data <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
head(data)
```

### Step 2: Tidy and Transform Data
Firstly, convert the `OCCUR_DATE` to the data date type, so that the month, year and day can be further extracted for analysis.

```{r}
#converting OCCUR_DATE to date data type
data <- data%>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, "%m/%d/%y"),
         YEAR = year(OCCUR_DATE))
```

Converting categorical variables to factor type not only improves analysis efficiency and accuracy but also ensures correct handling of these variables in statistical models and simplifies data operations. This is an important step in data cleaning and preprocessing, providing a solid foundation for subsequent data analysis and modeling.

```{r}
# filtering out observations fro Unknown victim age group
data <- data%>%
  filter(VIC_AGE_GROUP != 'UNKNOWN')

# Convert categorical variables to factor typ
data <- data%>%
  mutate(BORO = as.factor(BORO),
         PERP_AGE_GROUP = as.factor(PERP_AGE_GROUP),
         PERP_SEX = as.factor(PERP_SEX),
         PERP_RACE = as.factor(PERP_RACE),
         PERP_AGE_GROUP = as.factor(PERP_AGE_GROUP),
         VIC_AGE_GROUP = as.factor(VIC_AGE_GROUP),
         VIC_SEX = as.factor(VIC_SEX),
         VIC_RACE = as.factor(VIC_RACE)
         )
```

Key observations on data type conversion are:


* **BORO**: Represents the administrative districts of New York City, should be treated as a factor.
* **PERP_AGE_GROUP** and **VIC_AGE_GROUP**: Indicate the age groups of the perpetrators and victims, should be treated as a factor.
* **PERP_SEX** and **VIC_SEX**: Represent the gender of the perpetrators and victims, should be treated as a factor.
* **PERP_RACE** and **VIC_RACE**: Represent the race of the perpetrators and victims, should be treated as a factor.

Since I'd like to know which areas of New York City have seen higher rates of victimization of which populations in recent years, `VIC_AGE_GROUP`, `BORO`, and `VIC_SEX` are the three variables I picked up.

```{r}
# selecting variables of interest
subData <- data%>%
  select(VIC_RACE, BORO,  VIC_SEX)
# checking null values in selected data
colSums(is.na(subData))
```

From above output, it is pretty evident that there is no null values in the dataset which means the dataset is already cleaned.

```{r}
# summary of data
summary(subData)
```

The table shows that up to April 26, 2024, Brooklyn and Bronx have the highest number of shooting incidents, with 11,319 and 8,359 cases respectively. Among the victims, Black individuals have the highest representation, with 20,217 victims, followed by White Hispanics (4,275 victims) and Black Hispanics (2,791 victims). In terms of gender, male victims significantly outnumber female victims, with 25,737 males and 2,755 females.

### Step 3: Visualizations and Analysis

To further analyze whether conflict against ethnic minorities has intensified in recent years, I also correlate whether there is a correlation between the year of the epidemic and the ethnicity of the victims, i.e., whether there are any regions where the majority of incidents belong to victims of a particular age group.


```{r}
# Generate graphs comparing the ethnic composition and sex ratio of victims in different regions
ggplot(subData, aes(x = VIC_SEX, fill = VIC_RACE)) +
  geom_bar(position = "fill") +
  labs(title = "Victim Race Composition by Borough", y = "Proportion", x = "Borough") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ BORO, scales = "free_y")

```

The graph shows the proportion of different victim races by borough (Bronx, Brooklyn, Manhattan, Queens, and Staten Island) and gender (Female (F), Male (M), and Unknown (U)) in NYC. Here is the analysis based on the plot:

#### General Observations
1. **Black Victims**: 
   - The green bars represent Black victims, who make up the majority of victims in all boroughs for both male and female categories.
   - In the Bronx, Brooklyn, and Staten Island, the proportion of Black victims is particularly high, often exceeding 75% for both males and females.

2. **White Hispanic Victims**:
   - The pink bars represent White Hispanic victims, who form the second largest group in most boroughs.
   - In the Bronx, Brooklyn, and Queens, the White Hispanic victim proportion is significant but less than that of Black victims.

3. **Unknown Race Victims**:
   - The blue bars indicate victims whose race is unknown, making up a notable portion in Manhattan and Brooklyn for both males and females.

4. **Other Races**:
   - Other races, including Asian/Pacific Islander (orange), American Indian/Alaskan Native (red), and White (purple), have smaller proportions across all boroughs.
   - The Asian/Pacific Islander group has a visible presence in Queens and Manhattan but is less significant in other boroughs.

#### Gender-Specific Observations
1. **Male Victims**:
   - Across all boroughs, male victims (M) constitute the majority compared to female victims.
   - The race distribution among male victims follows the general trend with Black and White Hispanic victims being predominant.

2. **Female Victims**:
   - Female victims (F) also show a high proportion of Black victims across all boroughs.
   - The distribution for females is similar to that of males, with Black and White Hispanic victims being the most common.

3. **Unknown Gender Victims**:
   - There are very few or no victims recorded with unknown gender (U) in most boroughs except for a small number in Manhattan and Brooklyn.

#### Borough-Specific Observations
1. **Bronx**:
   - High proportion of Black victims.
   - Significant presence of White Hispanic victims.

2. **Brooklyn**:
   - Similar trend as Bronx with a high proportion of Black victims.
   - Noticeable presence of victims with unknown race.

3. **Manhattan**:
   - High proportion of Black victims, but also a significant number of unknown race victims.

4. **Queens**:
   - High proportion of Black victims.
   - Noticeable presence of Asian/Pacific Islander victims.

5. **Staten Island**:
   - High proportion of Black victims.
   - Proportion of White Hispanic victims is also significant.

### Conclusion
The analysis shows a clear predominance of Black victims in shooting incidents across all boroughs and genders, with the White Hispanic group being the second largest. There is a notable amount of data for unknown race and gender in some boroughs, which indicates areas where data completeness could be improved. This distribution highlights the racial disparities in shooting incidents in NYC and suggests a need for targeted intervention strategies.



```{r}
# Add year variable
data <- data %>%
  mutate(YEAR = as.numeric(format(OCCUR_DATE, "%Y")))

# Filter data for 2015-2019 and 2020-2024
data_2015_2019 <- data %>%
  filter(YEAR >= 2015 & YEAR <= 2019)
data_2020_2024 <- data %>%
  filter(YEAR >= 2020 & YEAR <= 2024)

# Combine data and add a period variable
data_combined <- bind_rows(
  data_2015_2019 %>% mutate(PERIOD = "2015-2019"),
  data_2020_2024 %>% mutate(PERIOD = "2020-2024")
)

# Generate graphs comparing the ethnic composition and sex ratio of victims in all areas over two time periods
ggplot(data_combined, aes(x = PERIOD, fill = VIC_RACE)) +
  geom_bar(position = "fill") +
  labs(title = "Victim Race Composition by Period", y = "Proportion", x = "Period") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ VIC_SEX, scales = "free_y")
```

### Conclusion

The second graph provides a focused view on the pandemic period (2020-2024) and reveals:

  - Consistent high proportions of Black victims across genders.
  - Significant presence of unknown race data in the unknown gender category.
  - Possible changes or continuities in victimization patterns during the pandemic years compared to previous years.
  - The insights gathered from the second graph complement the broader geographical analysis provided by the first graph, emphasizing the need to consider both temporal and geographical dimensions in understanding victimization patterns.

However, since previous tests have shown significant correlations between VIC_RACE and BORO, and between VIC_RACE and PERIOD, it was necessary to further test the independence between the three variables `VIC_RACE`, `BORO` and `VIC_SEX`.

```{R}
# Chi-square test for VIC_RACE and BORO
chi_square_test_race_boro <- chisq.test(table(data$VIC_RACE, data$BORO))
print(chi_square_test_race_boro)

# Chi-square test for VIC_RACE and VIC_SEX
chi_square_test_race_sex <- chisq.test(table(data$VIC_RACE, data$VIC_SEX))
print(chi_square_test_race_sex)

# Chi-square test for BORO and VIC_SEX
chi_square_test_boro_sex <- chisq.test(table(data$BORO, data$VIC_SEX))
print(chi_square_test_boro_sex)
```

These results indicate several significant statistical associations:

1. `VIC_RACE` and `BORO`:
   - X-squared = 2583.9, df = 24, p-value < 2.2e-16
   - This result shows a highly significant association between victim race (VIC_RACE) and the borough (BORO) where the crime occurred. The low p-value (< 0.05) indicates significant variation in the distribution of victim races across different boroughs.

2. `VIC_RACE` and `VIC_SEX`:
   - X-squared = 444.29, df = 12, p-value < 2.2e-16
   - This result demonstrates a significant association between victim race (VIC_RACE) and gender (VIC_SEX). The low p-value (< 0.05) suggests substantial differences in the distribution of victim races across different gender groups.

3. `BORO` and `VIC_SEX`:
   - X-squared = 17.866, df = 8, p-value = 0.02225
   - This result shows an association between the borough (BORO) where the crime occurred and the victim's gender (VIC_SEX). Although the p-value is relatively small (0.02225), it still indicates a statistically significant relationship between these two variables.

4. `VIC_RACE` and `PERIOD`:
   - X-squared = 78474, df = 6, p-value < 2.2e-16
   - This result indicates a highly significant association between victim race (VIC_RACE) and the time period (PERIOD) in which the crime occurred. The low p-value (< 0.05) suggests significant variation in the distribution of victim races across different time periods.

In summary, these findings demonstrate significant statistical associations between victim race, gender, geographical location, and the time period of crime occurrence. These associations are crucial for further research and policy-making considerations.


### Conclusion
Overall Summary
Both chi-square tests indicate that there are significant relationships between victim race and both the borough and the period. Specifically:

  - **Victim race distribution varies by borough:** Different boroughs have different proportions of victims from various racial groups.
  - **Victim race distribution varies by period:** The racial composition of victims has changed significantly between the periods compared, which could be related to socio-political factors, including the impact of the COVID-19 pandemic and changes in social dynamics.
  
This suggests that further investigation into the socio-economic and policy factors influencing these distributions could be valuable for understanding and addressing disparities in victimization across different racial groups and locations.

### Bias Identification
Given that these data are provided by the New York City Police Department, which has faced allegations of racial bias and unfair treatment towards minority groups, there may be inherent biases since the data collection is not independent. However, it's important to note that any assertions regarding bias cannot be definitively confirmed or validated, as I lack detailed knowledge of the inner workings of the New York City Police Department. My analysis is based on the available data, and I have made efforts to draw conclusions without making unfounded assumptions about biases present in this dataset.



